# Computational Environment and Data Ingestion

We begin by loading the necessary R libraries and reading the dataset.

```{r}
#| label: setup-libraries-data
#| echo: true
#| message: false
#| warning: false
#| results: 'hide'

# Initialize environment and load dependencies
librarian::shelf(
  tidyverse,
  tidymodels,
  kableExtra,
  patchwork,
  skimr,
  gridExtra,
  gtsummary,
  janitor,
  corrplot,
  sjPlot,
  scales,
  GGally,
  car,
  forcats,
  performance,
  glmmTMB,
  splines,
  mgcv,
  DHARMa,
  zoo,
  ggpubr,
  ggridges,
  caret,
  rstatix,
  Metrics,
  mice,
  missRanger,
  ranger,
  cocor,
  splines,
  multcompView,
  lmtest,
  aod,
  pROC
)

# Load the dataset
df <- read_csv("data/weatherAUS.csv")
source(here::here("utils.R"))
```

This section establishes the computational framework necessary for the analysis. We utilize the `librarian::shelf` function for robust dependency management, ensuring that all required packages are installed and loaded efficiently.

The analytical toolkit assembled here is comprehensive:

-   **Data Wrangling & Cleaning:** The `tidyverse` suite and `janitor` are employed for efficient data manipulation and cleaning.
-   **Advanced Statistical Modeling:** `glmmTMB` and `mgcv` are included to handle complex regression tasks, specifically zero-inflated models and Generalized Additive Models (GAMs), while `DHARMa` and `performance` are reserved for rigorous model diagnostics and residual analysis.
-   **Imputation & Machine Learning:** `missRanger` and `ranger` provide the framework for the Random Forest-based imputation strategy mentioned in the objectives.
-   **Visualization:** A combination of `patchwork`, `ggpubr`, and `ggridges` is loaded to generate publication-quality, multi-panel figures.

Finally, the primary dataset is ingested into the environment as `df` for immediate processing.


# Data Cleaning and Preprocessing

```{r}
#| label: data-cleaning
#| echo: true
#| message: false
#| warning: false

df_clean <- df %>%
  clean_names() %>%
  mutate(
    date = as.Date(date),
    month = as.factor(month(date)),
    day = as.factor(wday(date, label = TRUE))
  ) %>%
  # Remove records where the target variable (rainfall) is missing
  filter(!is.na(rainfall))


df_clean %>%
  head() %>%
  kable(caption = "Head of Cleaned Dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

df_clean %>%
  tail() %>%
  kable(caption = "Tail of Cleaned Dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

In this phase, we transform the raw dataset `df` into a consistent format suitable for analysis (`df_clean`). The preprocessing pipeline involves three primary operations:

1.  **Syntactic Standardization:** The `clean_names()` function is applied to convert all column headers to `snake_case`, ensuring consistent variable naming conventions across the dataset.
2.  **Temporal Feature Extraction:** The `date` column is cast to a proper Date object. From this, we derive two key categorical variables: `month` (to capture seasonal rainfall variations) and `day` (representing the day of the week). These derived features will be essential for identifying temporal patterns in precipitation.
3.  **Target Integrity:** We explicitly exclude observations where the target variable, `rainfall`, is missing (`NA`). Since our primary objective is to model rainfall occurrence and intensity, records without this ground truth provide no supervised learning value and are therefore removed.

The resulting tables (Head and Tail) confirm the successful standardization of variables and the existence of the newly engineered temporal features.


# Hybrid Imputation Strategy

```{r}
#| label: imputation-strategy
#| echo: true
#| eval: false

clean_and_impute_weather <- function(df) {
  # Configuration
  MAXGAP <- 5
  GHOST_THRESHOLD <- 0.90
  PMM_K <- 5
  MAXITER <- 4

  # Initial cleaning
  df <- df %>%
    clean_names() %>%
    mutate(
      date = as.Date(date),
      month = as.factor(month(date)),
      day = as.factor(wday(date, label = TRUE)),
      day_of_year = yday(date)
    ) %>%
    filter(!is.na(rainfall)) %>%
    select(-rain_tomorrow)

  # Create imputation flags

  df_flagged <- df %>%
    mutate(
      sunshine_imp_flagged = ifelse(is.na(sunshine), 1, 0),
      evap_imp_flagged = ifelse(is.na(evaporation), 1, 0),
      cloud3pm_imp_flagged = ifelse(is.na(cloud3pm), 1, 0),
      cloud9am_imp_flagged = ifelse(is.na(cloud9am), 1, 0)
    )

  # Temporal interpolation

  interp_vars <- c(
    "min_temp",
    "max_temp",
    "temp9am",
    "temp3pm",
    "pressure9am",
    "pressure3pm",
    "humidity9am",
    "humidity3pm"
  )

  df_interp <- df_flagged %>%
    group_by(location) %>%
    arrange(date, .by_group = TRUE) %>%
    mutate(across(
      all_of(interp_vars),
      ~ na.approx(., maxgap = MAXGAP, na.rm = FALSE, rule = 2)
    )) %>%
    ungroup()

  # Identify ghost sensors

  ghost_prone_vars <- c("sunshine", "evaporation", "cloud3pm", "cloud9am")

  ghost_pairs <- df_interp %>%
    select(location, all_of(ghost_prone_vars)) %>%
    pivot_longer(
      cols = all_of(ghost_prone_vars),
      names_to = "variable",
      values_to = "value"
    ) %>%
    group_by(location, variable) %>%
    summarise(
      miss_rate = mean(is.na(value)) * 100,
      .groups = "drop"
    ) %>%
    filter(miss_rate > (GHOST_THRESHOLD * 100)) %>%
    select(location, variable)

  cat(sprintf("  Found %d ghost sensor instances\n\n", nrow(ghost_pairs)))

  # missRanger imputation (first pass for general variables)

  imputation_data <- df_interp %>%
    mutate(
      sin_month = sin(2 * pi * as.numeric(month) / 12),
      cos_month = cos(2 * pi * as.numeric(month) / 12),
      sin_doy = sin(2 * pi * day_of_year / 365),
      cos_doy = cos(2 * pi * day_of_year / 365)
    )

  metadata_cols <- imputation_data %>% select(date)
  imputation_cols <- imputation_data %>% select(-date)

  imputed_data <- missRanger(
    data = imputation_cols,
    pmm.k = PMM_K,
    num.trees = 100,
    sample.fraction = 0.3,
    min.node.size = 10,
    seed = 123,
    verbose = 1,
    maxiter = MAXITER
  )

  df_imputed <- bind_cols(metadata_cols, imputed_data) %>%
    select(-starts_with("sin_"), -starts_with("cos_"))

  # Sanitize ghost sensors

  if (nrow(ghost_pairs) > 0) {
    ghost_map <- split(ghost_pairs$location, ghost_pairs$variable)

    df_imputed <- df_imputed %>%
      mutate(across(
        names(ghost_map),
        ~ replace(., location %in% ghost_map[[cur_column()]], NA)
      ))

    cat(sprintf("  Reverted %d instances to NA\n\n", nrow(ghost_pairs)))
  }

  # Targeted MICE with Random Forest for high-missingness variables

  init <- mice(df_imputed, maxit = 0)
  pred <- init$predictorMatrix
  meth <- init$method
  pred[,] <- 0

  # Set method to random forest for target variables
  meth["sunshine"] <- "rf"
  meth["evaporation"] <- "rf"
  meth["cloud9am"] <- "rf"
  meth["cloud3pm"] <- "rf"

  # Define predictors for each variable
  sun_predictors <- intersect(
    colnames(df_imputed),
    c("cloud9am", "cloud3pm", "max_temp", "humidity3pm", "location", "month")
  )

  evap_predictors <- intersect(
    colnames(df_imputed),
    c(
      "wind_gust_speed",
      "max_temp",
      "humidity3pm",
      "sunshine",
      "location",
      "month"
    )
  )

  cloud_predictors <- intersect(
    colnames(df_imputed),
    c("humidity9am", "humidity3pm", "pressure9am", "location", "month")
  )

  if ("sunshine" %in% rownames(pred)) {
    pred["sunshine", sun_predictors] <- 1
  }
  if ("evaporation" %in% rownames(pred)) {
    pred["evaporation", evap_predictors] <- 1
  }
  if ("cloud9am" %in% rownames(pred)) {
    pred["cloud9am", cloud_predictors] <- 1
  }
  if ("cloud3pm" %in% rownames(pred)) {
    pred["cloud3pm", cloud_predictors] <- 1
  }

  ignore_cols <- grep("_flagged$|^date$", colnames(pred), value = TRUE)
  pred[, ignore_cols] <- 0

  imp <- mice(
    df_imputed,
    method = meth,
    predictorMatrix = pred,
    m = 1,
    maxit = 5,
    seed = 123,
    printFlag = FALSE
  )

  df_final <- complete(imp)

  return(df_final)
}
```


```{r}
#| label: clean-and-save-imputed
#| message: false
#| warning: false
#| eval: false

df_final <- clean_and_impute_weather(df)
write_csv(df_final, "data/df_final.csv")
```

```{r}
#| label: load-imputed-data
#| echo: false
#| message: false
#| warning: false

# Load the pre-computed dataset to save rendering time
df_final <- read_csv("data/df_final.csv")
```

To address the missing data challenges identified in the EDA (specifically the high missingness in `sunshine` and `evaporation`), we implemented a robust, multi-stage hybrid imputation strategy designed to preserve the statistical properties of meteorological data.

**1. Informative Missingness Flags:** Before imputation, we generated binary flags (e.g., `sunshine_imp_flagged`) for variables with high missingness. This ensures that if the absence of data itself carries information (e.g., a broken sensor during a specific storm type), the model retains the capacity to learn from it.

**2. Temporal Interpolation (Time-Series Logic):** Recognizing that weather variables like temperature and pressure exhibit strong temporal autocorrelation, we first applied linear interpolation (`na.approx`) grouped by `Location`.

-   **Rationale:** If the temperature is $20^\circ$C on Monday and $22^\circ$C on Wednesday, it is physically sound to estimate Tuesday as $21^\circ$C rather than using a global average.
-   **Constraint:** This was limited to a `maxgap` of 5 days to prevent creating artificial data over long periods of sensor failure.

**3. Multivariate Imputation via Chained Random Forests:** For remaining gaps, particularly those in non-continuous variables like `cloud cover` or `wind direction`, we utilized the `missRanger` algorithm.

-   **Method:** This technique uses Random Forests to predict missing values based on all other available variables iteratively.
-   **Advantage:** Unlike mean imputation, which shrinks variance, `missRanger` with Predictive Mean Matching (PMM) preserves the original distribution and the complex non-linear correlations between weather features.

This comprehensive approach ensures `df_final` is complete, statistically sound, and ready for advanced modeling.

```{r}
#| label: save-clean-data
#| include: false

if (!dir.exists("data")) {
  dir.create("data")
}

saveRDS(df_clean, "data/df_clean.rds")
```
